---
title: "gov51-ps4"
format: html
editor: visual
---

## Gov 51 PSET 4

<https://github.com/sagepiekarski-ai/gov51-ps4.git>

### Sage Piekarski

#### 1 Setup and Exploration

```{r}
#1.1
rm(list=ls())
cat('\014')

#Install and load packages
if (!require(tidyverse)) install.packages("tidyverse"); library(tidyverse)
if (!require(estimatr)) install.packages("estimatr"); library(estimatr)
if (!require(modelsummary)) install.packages("modelsummary"); library(modelsummary)

card <- read_csv("Data/card1995.csv")
```

1.2 Describe the data

In the dataset there are 3010 observations. The unit of observations are individual workers.

```{r}
#1.3
ggplot(data = card, aes(x = educ))+
  geom_histogram(bins = 15, color = "white", fill = "cornflowerblue") +
  labs(title = "Distribution of Years of Completed Education",
       x = "Years of Completed Education (endogenous variable)",
       y = "count")
mean(card$educ, na.rm = TRUE)
```

1.3 Distribution of years of education

The mean is 13.263. Based on the histogram there appear to be large spikes at about 12 years of education and 16 years of education. Conceptually, it makes sense that there would be spikes here because 12 years of education is associated with completion of a high school degree, and 16 years would be consistent with completion of a bachelors degree. It makes sense that people would stop with education at these milestone points.

```{r}
#1.4
ggplot(data = card, aes(x=educ, y=lwage))+
  geom_point(alpha = 0.2)+
  geom_smooth(method = "lm", se = FALSE, color = "cornflowerblue")+
  labs(title = "Log Weekly Wage vs. Years of Completed Education",
       x = "Years of Completed Education (endogenous variable)",
       y = "Log Weekly Wage (outcome)")
```

1.4 The naïve relationship

Based on this plot, more education does appear to be associated with higher wages as can be seen by the positive slope of our linear regression line.

1.5 Why OLS might be misleading

The OLS regression of lwage on educ might not give us the causal effect of education. One reason why people with more education might earn more even if education itself had no effect is because people with more education, may be those with more financial resources to go to school for longer, and these financial resources also allow them to have more more connections and better job opportunities via a higher earning network. Another reason why people with more education might earn more even if education itself has no effect, is if people who get more education are fundamentally those that have higher performance abilities. As a result, these people are also more systematically likely to do better at work, have better job, and have higher earnings.

#### 2 The Instrument: College Proximity 

