---
title: "gov51-ps4"
format: html
editor: visual
---

## Gov 51 PSET 4

<https://github.com/sagepiekarski-ai/gov51-ps4.git>

### Sage Piekarski

#### 1 Setup and Exploration

```{r}
#1.1
rm(list=ls())
cat('\014')

#Install and load packages
if (!require(tidyverse)) install.packages("tidyverse"); library(tidyverse)
if (!require(estimatr)) install.packages("estimatr"); library(estimatr)
if (!require(modelsummary)) install.packages("modelsummary"); library(modelsummary)

card <- read_csv("Data/card1995.csv")
```

**1.2 Describe the data**

In the dataset there are 3010 observations. The unit of observations are individual workers.

```{r}
#1.3
ggplot(data = card, aes(x = educ))+
  geom_histogram(bins = 15, color = "white", fill = "cornflowerblue") +
  labs(title = "Distribution of Years of Completed Education",
       x = "Years of Completed Education (endogenous variable)",
       y = "count")
mean(card$educ, na.rm = TRUE)
```

**1.3 Distribution of years of education**

The mean is 13.263. Based on the histogram there appear to be large spikes at about 12 years of education and 16 years of education. Conceptually, it makes sense that there would be spikes here because 12 years of education is associated with completion of a high school degree, and 16 years would be consistent with completion of a bachelors degree. It makes sense that people would stop with education at these milestone points.

```{r}
#1.4
ggplot(data = card, aes(x=educ, y=lwage))+
  geom_point(alpha = 0.2)+
  geom_smooth(method = "lm", se = FALSE, color = "cornflowerblue")+
  labs(title = "Log Weekly Wage vs. Years of Completed Education",
       x = "Years of Completed Education (endogenous variable)",
       y = "Log Weekly Wage (outcome)")
```

**1.4 The naïve relationship**

Based on this plot, more education does appear to be associated with higher wages as can be seen by the positive slope of our linear regression line.

**1.5 Why OLS might be misleading**

The OLS regression of lwage on educ might not give us the causal effect of education. One reason why people with more education might earn more even if education itself had no effect is because people with more education, may be those with more financial resources to go to school for longer, and these financial resources also allow them to have more more connections and better job opportunities via a higher earning network. Another reason why people with more education might earn more even if education itself has no effect, is if people who get more education are fundamentally those that have higher performance abilities. As a result, these people are also more systematically likely to do better at work, have better job, and have higher earnings.

#### 2 The Instrument: College Proximity 

```{r}
mean(card$nearc4, na.rm=TRUE)
```

**2.1 What the instrument measures**

The variable nearc4 equals 1 if the respondent grew up near a 4-year college. The share of respondents in this sample that grew up near a 4-year college is 0.682 or about 68.2 people per 100 respondents (68.2%).

```{r}
#2.2
first_stage <- lm_robust(educ ~ nearc4 + black + smsa + south + exper + expersq + married,
                         data = card, se_type = "HC2")
first_stage
```

**2.2 Relevance: does proximity affect education?**

The coeﬀicient on nearc4 is 0.327 and its standard error is 0.081. The t-statistic is 4.061. This indicates that living near a 4-year college at age 14 was on average associated with the completion of an additional 0.327 years of education holding constant being black, lived in a metropolitan area, lived in the South, potential labor market experience (years), experience squared, and married. The positive sign makes sense because if one lives near a college as a child they likely experienced a reduced cost of education meaning it is easier and cheaper for them to go to college.

```{r}
F <- 4.0607488^2
F
```

**2.3 Checking instrument strength**

The t-statistic on nearc4 from the first-stage regression was 4.061. The first-stage F-statistic is 16.490. Our instrument is strong because 16.49 \> 10. Instrument strength matters for the reliability of the 2SLS estimate because the point of including instrumental vairables is to establish a causal relationship and if you include a weak instrument it leaves a lot of room for variability, imprecision and bias.

**2.4 Exclusion restriction**

The second condition for a valid instrument is the exclusion restriction: college proximity must affect wages only through its effect on education — not through any other channel. (a) The exclusion restriction in this example is that it must be true that once we control for education, nearc4 must only affect lwage through its affect on educ. (b) One way the exclusion restriction might be violated is that areas around colleges may systematically differ in terms of racial segregation or racial profiling. In this case, employers around colleges may differ in hiring discrimination than in other areas. Perhaps near these areas employers discriminate less throughout the hiring process, so if you are a person of color and live in this area you may have other wages than those that do no live near colleges.

```{r}
#2.5
card_summ <- card %>% 
  group_by(nearc4) %>% 
  summarise(mean_black = mean(black),
            mean_smsa = mean(smsa),
            mean_south = mean(south))
card_summ

```

2.5 Balance check

The groups do not appear to be balanced. If we look at the table there appear to be slightly less black people that lived near colleges, there appear to be a substantial amount more people that lived in a metropolitan area that lived near colleges, and there appear to be less people that lived in the south that lived near colleges. These imbalances tell us that there may be a lot of outside variables affecting wage. In our study black, smsa, and south are not randomly assigned and nearc4 is not as good as randomly assigned so exogeneity could be a threat to the validity of the instrument.

#### 3 The Wald Estimator 

```{r}
#3.1
reduced_form <- lm_robust(lwage ~ nearc4 + black + smsa + south + exper + expersq + married,
                          data = card, se_type = "HC2")
reduced_form
```

**3.1 Compute the reduced form**

The coeﬀicient on nearc4 is 0.041. This tells us that those that living near a college is on average associated with a 4.1% increase in wages, holding constant being black, lived in a metropolitan area, lived in the South, potential labor market experience (years), experience squared, and married.

```{r}
#3.2

fstage_coeff <- 0.327111413 
rform_coeff <- 0.040854001

wald_estimator <- rform_coeff/fstage_coeff
wald_estimator
```

**3.2 Calculate the Wald estimator by hand**

The Wald estimator is 0.125.

**3.3 Interpret the Wald estimator**

The wald estimate says that the effect of one additional year of education is associated with a with a change in outcome, wages, of +12.5%.

#### 4 OLS vs 2SLS

```{r}
#4.1
ols <- lm_robust(lwage ~ educ + black + smsa + south + exper + expersq + married,
                 data = card, se_type = "HC2")
ols
```

**4.1 OLS Estimate**

The coefficient on educ is 0.071

```{r}
#4.2
iv <- iv_robust(lwage ~ educ + black + smsa + south + exper + expersq + married | nearc4 + black + smsa + south + exper + expersq + married, data = card, se_type = "HC2")
iv
```

**4.2 2SLS Estimate**

The coeﬀicient on educ is 0.125. It does match the Wald estimate from Section 3.2.

```{r}
#4.3
modelsummary(list("OLS" = ols, "2SLS" = iv), coef_map = c("educ" = "Years of Education"), gof_map = c("nobs", "r.squared"))
```

**4.4 Why do OLS and 2SLS differ?**

\(a\) The coefficient of the 2SLS is greater.

\(b\) Card (1995) interprets the finding that 2SLS \> OLS as evidence of downward bias in OLS. One thing that could cause OLS to underestimate the return to education is measurement error. If we think about educ, it is an endogenous variable that measures years of completed education. If in fact this variable is measured incorrectly, it could indicate that education has a greater effect on wage than the model expected, this is attenuation bias.
(c) I believe the estimate from the 2SLS is more credible as a causal estimate of the effect of education on wages. This is because the OLS estimate is at risk of the omitted variable bias, as the included variables are all at the subjective disgression of the person running the model. With 2SLS, the model is able to
eliminate variation due to the near college instrumental variable.

```{r}
#4.5
ols_iq <- lm_robust(lwage ~ educ + IQ + black + smsa + south + exper + expersq + married,
                  data = card, se_type = "HC2")
ols_iq

twosls_iq <- iv_robust(lwage ~ educ + IQ + black + smsa + south + exper + expersq + married | nearc4 + IQ + black + smsa + south + exper + expersq + married, data = card, se_type = "HC2")
twosls_iq

0.125 - 0.109
```

**4.5 Adding IQ as a control**

\(a\) Including IQ changes the OLS coeﬀicient on educ by decreasing it to 0.068 from 0.071. This

\(b\) The 2SLS estimate decreases to 0.109 from 0.125 or by 0.016.

\(c\) These comparisons between part (a) and part (b) suggest that IQ is one source of bias that impacts the model as it causes both the OLS and 2SLS estimates to decrease. However, the OLS and 2SLS estimates still remain relatively different indicating there are other reasonings why the IV estimate is greater than OLS.

#### 5 Interpretation and LATE

**5.1 Who are the compliers?**
In this context, the compliers are individuals who would not have otherwise gone to college but decided to go to college because they happened to grow up near a 4-year college. For instance, these could be people that living near a college allowed for the costs associated with going to college to be cheaper so they could go to a college because they lived near it.

5.2 Write the LATE interpretation

Write a single careful sentence interpreting the 2SLS coeﬀicient as a LATE. Your sentence should specify:

The

\(i\) The causal effect being measured

\(ii\) For which specific group of people

\(iii\) What source of variation identifies it
